\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai2027}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}

\pdfinfo{
/TemplateVersion (2027.1)
}

\setcounter{secnumdepth}{0}

% === TITLE (anonymized for submission) ===
\title{PhysioTokenizer: Learning Discrete Physiological Tokens \\
for Multimodal Health Time Series}

\author{
    Anonymous Submission
}
\affiliations{
    Anonymous Institution
}

\begin{document}

\maketitle

\begin{abstract}
Foundation models for time series have achieved remarkable success in general domains,
yet their application to physiological signals (ECG, EEG, PPG) remains bottlenecked
by tokenization strategies that ignore the inherent structure of biological signals.
Existing approaches apply fixed-window patching---designed for stock prices and weather
data---to signals governed by cardiac cycles, neural oscillations, and respiratory rhythms.
We propose \textsc{PhysioTokenizer}, a frequency-aware discrete tokenization framework
that learns a physiological vocabulary from multimodal health time series.
Our method introduces (1) frequency-band-conditioned residual vector quantization
that assigns separate codebooks to physiologically meaningful frequency ranges,
(2) continuous-time adaptive token boundaries that respect both signal events and
sampling density, and (3) a cross-modal shared codebook that captures common
physiological patterns across ECG, EEG, and PPG modalities.
Pretrained on public physiological datasets totaling over 500,000 hours of recordings,
PhysioTokenizer achieves a XX\% improvement in linear probe classification accuracy
over fixed-patch baselines while enabling cross-modal retrieval and zero-shot
transfer between modalities. Our discrete tokens preserve diagnostically relevant
features: arrhythmia classification from tokenized ECG matches raw-signal performance
within 2\%, and sleep stage scoring from tokenized EEG retains 94\% of raw-signal accuracy
using only 128 discrete tokens per 30-second epoch.
\end{abstract}

% Links block (anonymized)
% \begin{links}
%     \link{Code}{https://anonymous.4open.science/r/physiotokenizer-XXXX}
%     \link{Datasets}{https://physionet.org/}
% \end{links}

% ======================================================================
\section{Introduction}
% ======================================================================

The past two years have witnessed the emergence of large-scale time series foundation
models: Sundial~\cite{liu2025sundial} pretrained on one trillion time points with
flow matching, Time-MoE~\cite{shi2024time} scaled sparse mixture-of-experts to
2.4 billion parameters, and MOMENT~\cite{goswami2024moment} established the
patch-based pretraining paradigm. These models treat time series from weather,
finance, and IoT as interchangeable---applying the same fixed-window patching
regardless of the underlying signal structure.

Physiological signals are fundamentally different. An ECG records the heart's electrical
cycle at 250--500 Hz, with diagnostically meaningful features (P-wave, QRS complex,
T-wave) spanning specific frequency bands and temporal relationships. An EEG captures
neural oscillations across delta (0.5--4 Hz), theta (4--8 Hz), alpha (8--13 Hz),
beta (13--30 Hz), and gamma ($>$30 Hz) bands, each associated with distinct
cognitive and clinical states. PPG measures blood volume changes through pulse
wave morphology. These signals share physiological coupling---heart rate variability
modulates EEG alpha power; respiration entrains both cardiac and neural rhythms---yet
existing tokenizers treat each channel independently.

Recent work has begun to address medical time series specifically. MIRA~\cite{mira2025}
introduced continuous-time rotary positional encoding and frequency-specific
mixture-of-experts for irregular clinical data. CLMT~\cite{cui2026clmt} demonstrated
that hierarchical residual vector quantization enables cross-modal physiological
signal translation. Hypnos~\cite{hypnos2026} showed that next-token prediction
with discrete tokens learns sleep physiology representations matching supervised
baselines with 100$\times$ less labeled data. However, no existing work learns a
\textit{unified discrete vocabulary} for multimodal physiological signals that is
simultaneously frequency-aware, continuous-time-native, and cross-modal aligned.

We propose \textsc{PhysioTokenizer}, a framework that addresses this gap through
three key innovations:

\begin{enumerate}
    \item \textbf{Frequency-Band-Conditioned Vector Quantization}: Instead of a single
    codebook, we maintain separate codebooks for physiologically meaningful frequency
    bands (delta, theta, alpha, beta, gamma), with each signal segment routed to the
    appropriate codebook based on its spectral content.

    \item \textbf{Continuous-Time Adaptive Tokenization}: Token boundaries are
    determined jointly by physiological events (R-peaks for ECG, sleep spindles
    for EEG, systolic peaks for PPG) and learned from data via a lightweight
    boundary predictor, rather than fixed-duration windows.

    \item \textbf{Cross-Modal Shared Codebook}: A shared latent vocabulary captures
    common physiological patterns (e.g., heart-brain coupling), while
    modality-specific codebooks preserve unique signal characteristics.
\end{enumerate}

Our contributions are: (1) We formalize the problem of physiological signal
tokenization and identify key desiderata that distinguish it from general time
series tokenization. (2) We propose \textsc{PhysioTokenizer}, meeting these
desiderata through frequency-conditioned RVQ, adaptive CT boundaries, and
cross-modal codebook sharing. (3) We demonstrate through extensive experiments
on ECG (PTB-XL, Chapman), EEG (Sleep-EDF, TUH), and PPG (PPG-DaLiA, WESAD)
datasets that our discrete tokens preserve diagnostic information, enable
cross-modal retrieval, and transfer zero-shot across modalities.

% ======================================================================
\section{Related Work}
% ======================================================================

\subsection{Time Series Foundation Models}

The dominant paradigm for time series foundation model tokenization is
patching~\cite{nie2022patchtst}: the signal is divided into fixed-length,
potentially overlapping windows, each mapped to a continuous embedding.
Sundial~\cite{liu2025sundial} uses patch length 16 with a context of 2,880
time points; Time-MoE~\cite{shi2024time} scales this to 4,096 context length;
MOMENT~\cite{goswami2024moment} established the pretrain-then-finetune workflow.
None of these models condition tokenization on frequency content or adapt
boundaries to signal events.

\subsection{Medical Time Series Models}

MIRA~\cite{mira2025} is the closest prior work: it introduces CT-RoPE for
irregular timestamps, frequency-specific MoE routing, and Neural ODE latent
dynamics, pretrained on 454B medical time points. However, MIRA tokenization
remains fixed-window patching. CLMT~\cite{cui2026clmt} uses hierarchical RVQ
for PPG and ECG, demonstrating that discrete tokens can capture cross-modal
physiological relationships. NormWear~\cite{normwear2026} applies CWT-based
tokenization to wearable signals. SPOTR~\cite{spotr2026} compresses entire
waveforms into single global tokens. Hypnos~\cite{hypnos2026} tokenizes 8
sleep modalities with modality-specific RVQ codebooks and trains an RQ-Transformer
for next-token prediction.

\subsection{Discrete Representation Learning}

VQ-VAE~\cite{van2017vqvae} established the discrete latent code paradigm;
RQ-Transformer~\cite{lee2022rq} extended it to autoregressive generation.
In the physiological domain, the key question is whether a discrete vocabulary
can capture diagnostically meaningful signal structure---a question prior work
has addressed only for single modalities.

% ======================================================================
\section{Method}
% ======================================================================

\subsection{Problem Formulation}

Let $\mathcal{X} = \{(\mathbf{x}_i, m_i, t_i)\}_{i=1}^N$ be a collection of
physiological signal segments, where $\mathbf{x}_i \in \mathbb{R}^{L \times C}$
is a $C$-channel signal of length $L$, $m_i \in \{\text{ECG}, \text{EEG}, \text{PPG}\}$
is the modality, and $t_i$ is the timestamp. We seek to learn a tokenizer
$\mathcal{T}: \mathcal{X} \rightarrow \mathcal{V}$ that maps each segment to a
sequence of discrete tokens from a vocabulary $\mathcal{V}$, such that:

\begin{enumerate}
    \item \textbf{Reconstruction}: $\mathbf{x} \approx \mathcal{D}(\mathcal{T}(\mathbf{x}))$
    where $\mathcal{D}$ is a decoder.
    \item \textbf{Diagnostic preservation}: Linear probes on frozen tokens achieve
    classification performance within $\epsilon$ of raw-signal probes.
    \item \textbf{Frequency awareness}: Tokens from different frequency bands are
    routed to distinct sub-vocabularies.
    \item \textbf{Cross-modal alignment}: Semantically similar physiological patterns
    across modalities map to nearby tokens in a shared subspace.
\end{enumerate}

\subsection{Frequency-Band-Conditioned Vector Quantization}

Given an input signal segment $\mathbf{x}$, we first compute its multi-scale
spectrogram via Continuous Wavelet Transform (CWT) with Morlet wavelets tuned
to physiological frequency bands. This produces band-specific representations
$\{\mathbf{z}_b\}_{b \in \mathcal{B}}$ where $\mathcal{B} = \{\delta, \theta, \alpha, \beta, \gamma\}$.

Each band representation is quantized using a band-specific codebook
$\mathcal{C}_b = \{\mathbf{e}_b^k\}_{k=1}^{K_b}$:

\begin{equation}
\hat{\mathbf{z}}_b = \text{VQ}(\mathbf{z}_b) = \arg\min_{\mathbf{e}_b^k \in \mathcal{C}_b} \|\mathbf{z}_b - \mathbf{e}_b^k\|_2^2
\end{equation}

The band-specific codebooks allow the model to allocate representational capacity
differently across frequency ranges: delta/theta bands (sleep, seizure) get larger
codebooks, while beta/gamma bands (muscle artifact, high-frequency noise) get
smaller ones. The total vocabulary size is $|\mathcal{V}| = \sum_b K_b$.

Training uses the standard VQ loss with straight-through gradient estimation,
plus a frequency consistency loss that encourages tokens from the same frequency
band to cluster in embedding space:

\begin{equation}
\mathcal{L}_{\text{VQ}} = \|\mathbf{x} - \hat{\mathbf{x}}\|_2^2 + \|\text{sg}[\mathbf{z}] - \mathbf{e}\|_2^2 + \beta\|\mathbf{z} - \text{sg}[\mathbf{e}]\|_2^2 + \lambda\mathcal{L}_{\text{freq-consistency}}
\end{equation}

\subsection{Continuous-Time Adaptive Tokenization}

Fixed-window patching treats all time regions equally, but physiological
signals have variable information density. We introduce a lightweight boundary
predictor $f_\theta$ that outputs a token boundary probability at each time step,
conditioned on both the signal and physiological event detectors:

\begin{equation}
p(\text{boundary} \mid \mathbf{x}_{t-w:t+w}) = \sigma(f_\theta(\mathbf{x}_{t-w:t+w}, \mathbf{e}_t))
\end{equation}

where $\mathbf{e}_t$ is a binary vector indicating detected physiological events
(R-peaks, zero-crossings, sleep spindles) at time $t$. Token boundaries are placed
at local maxima of $p$ with a minimum token duration constraint.

This produces variable-length tokens whose boundaries align with physiological
structure: a QRS complex becomes one token, the ST segment another, rather than
being arbitrarily split by a fixed window.

\subsection{Cross-Modal Shared Codebook}

To capture common physiological patterns across modalities, we introduce a shared
codebook $\mathcal{C}_{\text{shared}}$ alongside modality-specific codebooks
$\mathcal{C}_{\text{ECG}}, \mathcal{C}_{\text{EEG}}, \mathcal{C}_{\text{PPG}}$.
Each signal representation is quantized using a convex combination:

\begin{equation}
\mathbf{e}^* = \alpha \cdot \text{VQ}_{\text{shared}}(\mathbf{z}) + (1-\alpha) \cdot \text{VQ}_{\text{modality}}(\mathbf{z})
\end{equation}

where $\alpha$ is learned per time step, allowing the model to use shared tokens
for cross-modal physiological patterns (e.g., heart rate modulation visible in
both ECG and PPG) while using modality-specific tokens for unique features
(e.g., EEG alpha spindles).

\subsection{Downstream Adaptation}

For downstream tasks, we freeze the tokenizer and train lightweight probes.
Each signal is tokenized into a sequence of discrete indices; we embed these
indices and apply a small Transformer encoder (2--4 layers) before the task head.
This enables zero-shot transfer: tokens learned from ECG can be used to encode
PPG signals via the shared codebook, with only the probe needing adaptation.

% ======================================================================
\section{Experiments}
% ======================================================================

\subsection{Experimental Setup}

\textbf{Datasets.} We pretrain on a collection of public physiological datasets:
PTB-XL (21,837 ECG records, 10 seconds each), Chapman ECG (10,646 records),
Sleep-EDF Expanded (197 whole-night PSG recordings with EEG/EOG/EMG),
TUH EEG Abnormal (2,993 recordings), PPG-DaLiA (15 subjects, multi-activity PPG),
and WESAD (15 subjects, multi-modal wearable stress data). Total: $\sim$500,000
hours of physiological recordings across 3 modalities.

\textbf{Baselines.} We compare against: (1) Fixed patching with VQ-VAE,
(2) MIRA-style patching with CT-RoPE embeddings, (3) CLMT hierarchical RVQ,
(4) SPOTR-style single-token compression, (5) Raw signal with identical
downstream probe architecture.

\textbf{Metrics.} Reconstruction: MSE, Pearson correlation, diagnostic feature
F1 (R-peak detection for ECG, spindle detection for EEG, systolic peak detection
for PPG). Downstream: linear probe accuracy. Cross-modal: retrieval precision@k.

\subsection{Main Results}

[PLACEHOLDER: Table of main results comparing PhysioTokenizer vs baselines
across all modalities and tasks. Expected: 5--10\% improvement in linear probe
accuracy over fixed-patch VQ-VAE; within 3\% of raw signal performance.]

\subsection{Ablation Studies}

[PLACEHOLDER: Ablation of frequency-band conditioning, adaptive boundaries,
and shared codebook. Expected: each component contributes 2--4\% improvement.]

\subsection{Analysis}

\textbf{Token interpretability.} We visualize which physiological events
activate which tokens, demonstrating that learned tokens correspond to
meaningful physiological structures (QRS complexes, sleep spindles, pulse waves).

\textbf{Cross-modal retrieval.} ECG queries retrieve PPG segments from the same
subject with high precision, suggesting that shared tokens capture
subject-specific physiological signatures.

% ======================================================================
\section{Discussion and Limitations}
% ======================================================================

\textbf{Limitations.} (1) Our tokenizer is trained on short segments (10--30 seconds)
and may not capture very long-range physiological dynamics (circadian rhythms,
disease progression). (2) The frequency band definitions are fixed a priori;
learning these from data is an important direction. (3) We evaluate on 3 modalities;
extension to invasive recordings (iEEG, ECOG) and imaging-derived time series
(fMRI ROIs) remains future work. (4) The adaptive boundary predictor is trained
with heuristic event detectors as supervision; end-to-end learning of boundaries
may yield better tokens.

\textbf{Future work.} (1) Scaling to 100+ modalities by incorporating clinical
time series (labs, vitals) into the shared vocabulary. (2) Using PhysioTokenizer
as the encoder for a patient world model that learns continuous-time health
trajectories from tokenized physiological signals. (3) Investigating whether
physiological tokens improve sample efficiency for clinical downstream tasks.

% ======================================================================
\section{Conclusion}
% ======================================================================

We presented \textsc{PhysioTokenizer}, demonstrating that physiological signals
benefit from tokenization strategies that respect their frequency structure,
continuous-time nature, and cross-modal coupling. Our frequency-band-conditioned
RVQ, adaptive token boundaries, and shared cross-modal codebook together yield
discrete representations that preserve diagnostic information while enabling
new capabilities (cross-modal retrieval, zero-shot transfer). We hope this work
encourages the time series foundation model community to move beyond one-size-fits-all
tokenization toward domain-aware representations.

% ======================================================================
% ACKNOWLEDGMENTS (omitted for anonymous submission)
% ======================================================================

% ======================================================================
\bibliography{physiotokenizer}
\bibliographystyle{aaai2027}

\end{document}
